Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-16 paper/paper.tex
finished draft of last Results section

Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any of the sequences from the Huanan Seafood Market (Figure~\ref{fig:tree_with_deleted}).
To visualize these sequences in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees rooted to put the putative progenitor at each of the three sequences with highest identity to the bat coronavirus outgroup (Figure~\ref{fig:tree_with_deleted}).

It is immediately apparent that the discrepancy between outgroup rooting and epidemiological data is substantially alleviated by adding the deleted sequences and properly annotating the Wuhan infections that were sequenced in Guangdong.

>>>>

Several of the early epidemic sequences from the deleted project and all of the sequences from patients infected in Wuhan but sequenced in Guangdong were more similar to the bat coronavirus outgroup than any sequences from the Huanan Seafood Market across the region in which all were fully sequenced (Figure~\ref{fig:deltadist_jitter}).
To place this observation in a phylogenetic context, I added the deleted sequences at compatible locations to the full-genome phylogenetic trees described in the previous section (Figure~\ref{fig:tree_with_deleted}).

It is immediately apparent that the discrepancy between outgroup rooting and epidemiological evidence that Wuhan was the origin of SARS-CoV-2 is alleviated by adding the deleted sequences and annotating Wuhan infections sequenced in Guangdong (Figure~\ref{fig:tree_with_deleted}).
The middle tree in Figure~\ref{fig:tree_with_deleted} now looks highly plausible, as half its progenitor node is derived from early Wuhan infections, which is more than any other equivalently large node.
The first known sequence identical to this putative progenitor (Guangdong/HKU-SZ-002/2020) is from a patient who developed symptoms on January 4 while visiting Wuhan~\citep{chan2020familial}.
In this scenario, the progenitor has three mutations towards the bat coronavirus outgroup relative to Wuhan-Hu-1 (C8782T, T28144C, and C29095T), and two mutations relative to proCoV2 (T18060C away from the outgroup and C29095T towards the outgroup).
The leftmost tree in Figure~\ref{fig:tree_with_deleted}, which has a putative progenitor identical to proCoV2~\citep{kumar2021evolutionary} also looks plausible, with some weight from Wuhan sequences.
Note that although three deleted early epidemic sequences are compatible with this proCoV2 progenitor, the defining C18060T mutation is not covered in the sequenced region---so further sequencing of these samples would be informative.
The rightmost tree in Figure~\ref{fig:tree_with_deleted} looks less plausible, as it has almost no weight from Wuhan and the first sequence identical to its progenitor was not collected until January 24.

We can also qualitatively evaluate the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the sample principle employed by \citet{worobey2020emergence} to help evaluate scenarios for the emergence of SARS-CoV-2 in Europe and North America: namely that during an exponentially growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
This principle is especially likely to hold for the scenarios represented in Figure~\ref{fig:tree_with_deleted}, where the are multiple observed representatives of each putative progenitor which each have the potential to yield descendants with new mutations.
Using this qualitative idea, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
I acknowledge that these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in more qualitative reasoning when there are valid concerns about the nature of the underlying data.
